---
title: "Simulating Likert item responses in R"
author: "Marko Lalovic"
date: "Last Updated: `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating Likert item responses in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "../man/figures/get_started/",
  dev = "svg")
```

This is a quick overview of the responsesR package:

* [Generating random item responses](#generating-random-responses)

* [Correlations between latent variables](#adjusting-correlations)

* [Parameters of the latent distribution](#adjusting-parameters)

    - [Mean](#adjusting-the-mean)
    
    - [Standard deviation](#adjusting-sd)
    
    - [Skewness](#adjusting-skewness)
    
* [Estimation of latent parameters](#estimation-of-parameters)

* [Reconstruction of latent parameters](#resonstruction-of-parameters)


## Generating random item responses {#generating-random-responses}
First, load the package responsesR, and set seed for reproducibility:
```{r}
library(responsesR) 
set.seed(12345)
```

To generate `n = 10` responses with `K = 5` response categories (on a 5-point Likert scale), use:
```{r}
get_responses(n = 10, K = 5)
```

To generate `n = 100` responses to multiple Likert items on different scales (for example K = 4, 5, 6), use:
```{r}
data <- get_responses(n = 100, K = c(4, 5, 6))
```

The result is an array of simulated responses, where columns represent responses to Likert items:
```{r}
head(data)
```

The corresponding bar plots are shown below:
```{r random_responses_barplots, fig.height=3, fig.width=10, fig.ext="svg"}
par(mfrow=c(1, 3))
for(i in 1:3) {
  barplot(table(data[, i]))
  title(paste("Y", i, sep = ""))
}
```

## Correlations between latent variables {#adjusting-correlations}
To set a specific correlation between latent variables, such as 0.5, set `R = 0.5`:
```{r}
data <- get_responses(n = 1000, K = c(4, 5, 6), R = 0.5)
cor(data)
```

Alternatively, provide a correlation matrix. For instance, create a 3 by 3 correlation matrix:
```{r}
R <- c(1.00, -0.63, -0.39, -0.63, 1.00, 0.41, -0.39, 0.41, 1.00)
R <- matrix(R, nrow=3)
R
```

Then, use it to generate responses to Likert items with a latent correlation matrix R:
```{r}
data <- get_responses(n = 1000, K = c(4, 5, 6), R = R)
cor(data)
```

To use a random correlation matrix, set `R = "random"`:
```{r}
data <- get_responses(n = 1000, K = c(4, 5, 6), R = "random")
cor(data)
```

## Parameters of the latent distribution {#adjusting-parameters}
By default, the function `get_responses` generates symmetrically distributed responses. To introduce asymmetry and alter the properties of hypothetical survey respondents, adjust the mean `mu`, standard deviation `sd`, and skewness `gamma1` of the latent distribution.

### Mean {#adjusting-the-mean}
To generate responses from latent distributions with different means, you can set the means using the `mu` parameter. For example, to set the means of latent distributions to -1, 0, and 1, execute:
```{r}
data <- get_responses(n = 1000, mu = c(-1, 0, 1))
```

The generated responses and the corresponding densities of latent distributions are displayed below.
```{r adjusting_the_mean, fig.height=6, fig.width=10, fig.ext="svg"}
par(mfrow=c(2, 3))
for(i in 1:3) {
  barplot(table(data[, i]))
  title(paste("Y", i, sep = ""))
}
x <- seq(-5, 5, length = 1000)
y <- lapply(c(-1, 0, 1), function(mu) dnorm(x, mean = mu))
for(i in 1:3) {
  plot(x, y[[i]], type="l", lwd = 2, xlab = "", ylab = "")
  title(paste("X", i, sep = ""))
}
```

### Standard deviation {#adjusting-sd}
You can adjust the standard deviations of latent distributions using the `sd` parameter. For instance, specifying `sd = c(0.5, 1, 1.5)`:
```{r}
data <- get_responses(n = 1000, sd = c(0.5, 1, 1.5))
```

This generates the following responses with corresponding densities of latent distributions depicted below:
```{r adjusting_the_sd, fig.height=6, fig.width=10, fig.ext="svg"}
par(mfrow=c(2, 3))
for(i in 1:3) {
  barplot(table(data[, i]))
  title(paste("Y", i, sep = ""))
}
y <- lapply(c(0.5, 1, 1.5), function(s) dnorm(x, sd = s))
for(i in 1:3) {
  plot(x, y[[i]], type="l", lwd = 2, xlab = "", ylab = "")
  title(paste("X", i, sep = ""))
}
```

### Skewness {#adjusting-skewness}
You can adjust the skewness of latent distributions using the `gamma1` parameter. For instance, specifying `gamma1 = c(-0.5, 0, 0.5)` sets the skewness of latent distributions to -0.5, 0, and 0.5:
```{r}
data <- get_responses(n = 10^6, gamma1 = c(-0.5, 0, 0.5))
```

The resulting responses with corresponding densities of latent distributions are displayed below:
```{r adjusting_the_skewness, fig.height=6, fig.width=10, fig.ext="svg"}
par(mfrow=c(2, 3))
for(i in 1:3) {
  barplot(table(data[, i]))
  title(paste("Y", i, sep = ""))
}
dp <- apply(cbind(rep(0, 3), rep(1, 3), c(-0.5, 0, 0.5)), 1, 
            function(x_row) { convert_params(as.vector(x_row)) })
x <- seq(-3, 3, length = 1000)
y <- lapply(1:3, function(i) { 
  d_skew_normal(x, dp[["xi", i]], dp[["omega", i]], dp[["alpha", i]]) 
})
for(i in 1:3) {
  plot(x, y[[i]], type="l", lwd = 2, xlab = "", ylab = "")
  title(paste("X", i, sep = ""))
}
```

The `gamma1` parameter controls the skewness of the latent distribution. The value of `gamma1` is limited to slightly less than the interval (-1, 1):
$$
-0.95 \leq \text{gamma1} \leq 0.95
$$

A positive value makes the distribution right-skewed, while a negative value makes it left-skewed.

## Estimation of latent parameters {#estimation-of-parameters}
In order to replicate existing survey data, you can estimate the parameters of latent distributions using the function `estimate_parameters`.

As an illustration, let's generate 1000 responses with an underlying normal distribution having `mu = -0.5` and `sd = 0.5`, and assuming there are `K = 5` response categories. Note that not all possible categories may have responses. In this example, `n`, `mu`, and `sd` are chosen such that no responses have a value of 5:
```{r estimation_of_parameters_example, fig.height=4, fig.width=4, fig.ext="svg"}
set.seed(12345)
data <- get_responses(n = 1000, mu = -0.5, sd = 0.5, K = 5)
par(mfrow=c(1, 1))
barplot(table(data))
```

In addition to the responses, the `estimate_parameters` function requires specifying the number of possible response categories `K` to return the estimates:
```{r}
estimate_parameters(data, K = 5)
```

You can also provide responses to multiple items. For instance, if we generate responses to three items with different latent parameter values:
```{r estimation_of_parameters_for_multiple_items, fig.height=4, fig.width=10, fig.ext="svg"}
data <- get_responses(n = 1000,
                      mu = c(1, 2, 3),
                      sd = c(0.5, 1, 1.5),
                      K = c(5, 6, 7))
par(mfrow=c(1, 3))
for(i in 1:3) {
  barplot(table(data[, i]))
  title(paste("Y", i, sep = ""))
}
```

The `estimate_parameters` function returns a table with estimates for each item:
```{r}
estimate_parameters(data, K = c(5, 6, 7))
```

To ensure good results, it is crucial to provide a large enough sample of responses, say `n > 200`. Otherwise, the estimated values may vary and deviate significantly from actual values. For instance, if we generate 100 observations from the same underlying latent distribution:
```{r estimation_of_parameters_small_sample, fig.height=4, fig.width=10, fig.ext="svg"}
data <- get_responses(n = 100,
                      mu = c(1, 2, 3),
                      sd = c(0.5, 1, 1.5),
                      K = c(5, 6, 7))
par(mfrow=c(1, 3))
for(i in 1:3) {
  barplot(table(data[, i]))
  title(paste("Y", i, sep = ""))
}
```
Then, although the latent distributions are the same as before, the estimated values differ and may not align with the actual latent parameters:
```{r}
estimate_parameters(data, K = c(5, 6, 7))
```

## Reconstruction of latent parameters {#resonstruction-of-parameters}
You can also reconstruct the parameter values of the latent distribution based on the discrete probability distribution over possible response categories 1 through K. For instance, assuming the latent variable follows a normal distribution with 5 response categories, the function `estimate_mu_sd` can accurately recover the parameters `mu` and `sd`:
```{r}
pk <- list("1" = 0.313, "2" = 0.579, "3" = 0.105, "4" = 0.003)
estimate_mu_sd(pk = pk, K = 5)
```

Here, `pk` can also represent the vector of proportions of responses across categories 1 through K:
```
   pk = (number of responses with value k) / (the number of all responses)
```
Additionally, assuming the latent variable follows a skew-normal distribution, you can provide the value of `gamma1`:
```{r}
pk <- list("1" = 0.564, "2" = 0.370, "3" = 0.063, "4" = 0.004)
estimate_mu_sd(pk = pk, K = 5, gamma1 = 0.5)
```
